---
title: "Make Countries Move"
author: "Theis Daniel Jensen"
date: "19/3/2026"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

#  Explore global development with R
In this exercise, you will load a filtered `gapminder` dataset -  with a subset of data on global development from 1952 - 2007 in increments of 5 years - to capture the period between the Second World War and the Global Financial Crisis. 

**Your task: Explore the data and visualise it in both static and animated ways, providing answers and solutions to 7 questions/tasks within this script.** 

- **Submit to Github this script with solutions and its knitted version.**  
- **Submit to Brightspace a document with the (1) URL of this script and its knitted version in Github (2) answer of 250 words and a (3) visualisation of the Final Question**

## Get the necessary packages
First, start with installing and activating the relevant packages `tidyverse`, `gganimate`, and `gapminder` if you do not have them already. 
*Pay attention to what warning messages you get when installing `gganimate`, as your computer might need other packages than `gifski` and `av`*

```{r libraries, echo = TRUE}
# install.packages("gganimate")
# install.packages("gifski")
# install.packages("av")
# install.packages("gapminder")
library(tidyverse)
library(gganimate)
library(gifski)
library(av)
library(gapminder)
```

## Look at the data and tackle the tasks
First, see which specific years are actually represented in the dataset and what variables are being recorded for each country. Note that when you run the cell below, Rmarkdown will give you two results - one for each line - that you can flip between.
```{r}
str(gapminder)
unique(gapminder$year)
?gapminder
head(gapminder)
```

The dataset contains information on each country in the sampled year, its continent, life expectancy, population, and GDP per capita.

Let's plot all the countries in 1952.
```{r 1952}
library(scales)
theme_set(theme_bw())  # set theme to white background for better visibility
options(scipen = 999)

ggplot(subset(gapminder, year == 1952), aes(gdpPercap, lifeExp, size = pop/1000000)) +
  geom_point(aes(color = continent)) +
  scale_x_log10()+
  ggtitle("world in 1952") +
  labs(x ="GDP per capita", y = "Life expectancy")+
  theme(text = element_text(size=18))
gapminder %>%
  filter(year == 1952) %>%
  slice_max(gdpPercap, n=1)
 
```

...

We see an interesting spread with an outlier to the right. Explore who it is so you can answer question 2 below!


Next, you can generate a similar plot for 2007 and compare the differences
```{r 2007}
options(scipen = 999)

ggplot(subset(gapminder, year == 2007), aes(gdpPercap, lifeExp, size = pop/1000000)) +
  geom_point(aes(color = continent)) +
  scale_x_log10()+
  ggtitle("world in 2007") +
  labs(x ="GDP per capita", y = "Life expectancy")+
  theme(text = element_text(size=18))

gapminder %>%
  filter(year == 2007) %>%
  slice_max(gdpPercap, n=5)
```

...

The black bubbles are a bit hard to read, the comparison would be easier with a bit more visual differentiation.

### Questions for the static figures:

1. *Answer: why does it make sense to have a log10 scale (`scale_x_log10()`) on the x axis?* (hint: try to comment it out and observe the result)

It makes sense to have log10scale in the function, because the gdpPercapita in different countries varies from from under 1000 to over 100.000. By using the log10scale, the spread of the gdpPercapita in the different countries will be compressed and the graphs patterns will be easier to see


2. *Answer: In Figure 1: Who is the outlier (the richest country in 1952) far right on the x axis?* 

By using the code 
gapminder %>% 
filter(year == 1952) %>% 
slice_max(gdpPercap, n = 1),
 we obtain a table showing the country with the highest GDP per capita in 1952. The result is Kuwait, located in Asia, with a GDP per capita of 108,382.4


3. *Fix Figures 1 and 2: Differentiate the **continents** by color, and fix the axis labels and units to be more legible (**Hint**: the 2.50e+08 is so called "scientific notation". You want to eliminate it.)*

The commands to fixing the color, axis labels and units are these:

Color = geom_point(aes(color=continent))
Axis labels = ggtitle("World in 1952") +
  labs(x= "GDP per capita",
          y= "Life expectancy") + 
  theme(text = element_text(size=18))
Units = options(scipen=999)


4. *Answer: What are the five richest countries in the world in 2007?*

By using the code 
gapminder %>% 
filter(year == 2007) %>% 
slice_max(gdpPercap, n = 5),
 we obtain a table showing the five countries with the highest GDP per capita in 2007. The result from nr 1 to 5 Norway, Kuwait, Singapore, United States and Ireland. .





##  Make it move!

The comparison would be easier if we had the two graphs together, animated. We have a lovely tool in R to do this: the `gganimate` package. Beware that there may be other packages your operating system needs in order to glue interim images into an animation or video. Read the messages when installing the package.

Also, there are *two* ways of animating the gapminder ggplot.

### Option 1: Animate using transition_states() 

The first step is to create the object-to-be-animated
```{r anim1 start}
anim <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop/1000000)) +
  geom_point(aes(color = continent)) +
  scale_x_log10()+
  ggtitle("world from 1952 - 2007") +
  labs(x ="GDP per capita", y = "Life expectancy")+
  theme(text = element_text(size=18))
   # convert x to log scale
anim
```

...

This plot collates all the points across time. The next step is to split it into years and animate it. This may take some time, depending on the processing power of your computer (and other things you are asking it to do). Beware that the animation might appear in the bottom right 'Viewer' pane, not in this rmd preview. You need to `knit` the document to get the visual inside an html file.

```{r anim1}
anim + transition_states(year, 
                      transition_length = 1,
                      state_length = 1) +
  ggtitle("world in:{closest_state}")

```
...

Notice how the animation moves jerkily, 'jumping' from one year to the next 12 times in total. This is a bit clunky, which is why it's good we have another option. 


### Option 2 Animate using transition_time()
This option smooths the transition between different 'frames', because it interpolates and adds transitional years where there are gaps in the timeseries data.

```{r anim2}
anim2 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop/1000000)) +
  geom_point(aes(color = continent)) +
  scale_x_log10() + 
     ggtitle("world from 1952 - 2007") +
  labs(x ="GDP per capita", y = "Life expectancy")+
    # convert x to log scale
  transition_time(year)+
  ggtitle("world in: {frame_time}")
anim2
```

The much smoother movement in Option 2 will be much more noticeable if you add a title to the chart, that will page through the years corresponding to each frame.

Now, choose one of the animation options and get it to work. You may need to troubleshoot your installation of `gganimate` and other packages 

### Tasks for the animations:

5. *Can you add a title to one or both of the animations above that will change* 
 *in sync with the animation?* *(**Hint**: search labeling for `transition_states()` and `transition_time()` functions respectively)*

Yes, you can with this function:
anim + transition_states(year, 
                      transition_length = 1,
                      state_length = 1)+
  ggtitle("World in {closest_state}")


6. *Can you made the axes' labels and units more readable? Consider expanding the abbreviated labels as well as the scientific notation in the legend and x axis to whole numbers. Also, differentiate the countries from different continents by color*

By using the code options(scipen = 999), you get much more readable numbers on the x-axis, as this code converts the values into whole numbers instead of using scientific notation with letters and plus signs.

To divide the countries into different colors based on which continent they belong to, you can use the code geom_point(aes(color = continent)). This provides a better overview of which countries belong to the different continents.



### Final Question
7. **Is the world a better place today than it was in the year you were born?** *Answer this question using the gapminder data.  Define better either as more prosperous, more free, more healthy, or suggest another measure that you can get from gapminder. Submit a 250 word answer with an illustration to Brightspace. Include a URL in your Brightspace submission that links to the coded solutions in Github. [Hint: if you wish to have more data than is in the filtered gapminder, you can load either the `gapminder_unfiltered` dataset or download more historical data at https://www.gapminder.org/data/ ]*

The gapminder data used in this are only available to the year 2007. This is the reason as to why we are comparing the data from 2002 to the data from 2007. There are several methods to search for data that can answer the question “Is the world a better place today than it was in the year you were born?” We have chosen this method by doing this command:
gapminder_unfiltered%>%
 filter(year %in% c(2002,2007)) %>%
  group_by(year)%>%
  summarise(weighted_lifeExp=weighted.mean(lifeExp,pop))
This function calculates the global life expectancy, where every country is weighted with its population for the years 2002 and 2007. 
The results are:
2002: 67.7
2007: 68.8
Even after 5 years, the life expectancy on a global level has risen by 1.2 years, which indicates that factors that impact life expectancy such as, child mortality, welfare, medicine has improved and the average people are therefore more healthy 
another useful function is this: 
gapminder_unfiltered %>%
  filter(year %in% c(2002, 2007)) %>%
  group_by(year) %>%
  summarise(weighted_gdpPercap = weighted.mean(gdpPercap, pop))
This code calculates the global populated-weighted GDP per capita for the years 2002 and 2007
The results are: 
2002: 7998.131
2007: 9353.027
Again, after 5 years the global populated-weighted GDP per capita has risen by 1354,89, which says that the average people have a higher income in the year and therefore are more prosperous.
With the use of these codes, the conclusion is that the world has become a better place from the year 2002 to 2007. 


```{r}
gapminder_unfiltered%>%
  filter(year %in% c(2002,2007))%>%
  group_by(year)%>%
  summarise(weighted_lifeExp = weighted.mean(lifeExp, pop))

gapminder_unfiltered%>%
  filter(year %in% c(2002,2007))%>%
  group_by(year)%>%
  summarise(weighted_gdpPercap = weighted.mean(gdpPercap, pop))

#illustration

compare_year <- gapminder_unfiltered %>%
  filter(year %in% c(2002, 2007)) %>%
  group_by(year) %>%
  summarise(weighted_lifeExp = weighted.mean(lifeExp,pop))
ggplot(compare_year,aes(factor(year),weighted_lifeExp))+
  geom_col(fill = "steelblue")+
  labs(
    title = "Global life expectancy in 2002 and 2007",
    x = "Year",
    y = "life expectancy"
  ) +
  coord_cartesian(ylim = c(65, 70))



```

